---
title: "Thresholded ORA Summary"
author: "Feifei Li"
date: "`r Sys.Date()`"
output:
    html_document:
        df_print: paged
---

```{r, include=FALSE}
source("../scripts/diff_expr_qlf.R", chdir = TRUE, local = knitr::knit_global())
source("../utils/gprofiler_wrapper.R",
       chdir = TRUE,
       local = knitr::knit_global())
```

Three thresholded ORA's on all the top-hit differentially expressed genes, the up-regulated top-hits, and the down-regulated top-hots from the **QLF** result, were performed respectively using [g:Profiler](https://cran.r-project.org/web/packages/gprofiler2/vignettes/gprofiler2.html)[@gprofiler].


Since this is a pathway analysis, we are only interested in data sources for biological pathways. Hence, **Reactome**, **Wikipathways**, and **KEGG** were used as our annotation data sources. We will also include **GO:Biological process (GO:BP)** as our data source. Although a biological process is different from a pathway; however, pathways are considered to collectively participate in biological processes. Therefore, GO:BP was too included.
The databases used by the current release of g:Profiler are *Ensembl 102*, which is also the current version of GO database, and *Ensembl Genomes 49*.
The currect release of Reactome used is *version 75*. Wikipathways is a comunity-maintained database and is being constantly updated. The current release of KEGG used is *97.0*.

The p-value threshold for the significance of the gene-set enrichment of the Fisher's exact test calculated by g:Profiler was set to $0.05$. The numbers of genesets that were returned with this threshold by each query would be presented as a summary below. Benjamini–Hochberg FDR for p-value adjustment method was chosen, and IEA-annotated results (Inferred from Electronic Annotation) were excluded since the human-curated results are desirable for our first analysis.


We first performed a thresholded ORA on *all* the QLF significant genes. Below is a table summary of its g:Profiler query result:
```{r gost_all, message=FALSE, warning=FALSE, echo=FALSE}
gostres_tophits <- gostQuery(genelist = tophits_qlf)
knitr::kable(summariseGost(gostres = gostres_tophits))
```


An interactive Manhattan plot combining the results of different annotation sources:
```{r, warning=FALSE, message=FALSE, echo=FALSE}
gprofiler2::gostplot(gostres = gostres_tophits)
```

As for the detailed query result, however, there were too many pathways to check. Since GO terms are propagated,
there might be terms that are too board and not so informative.
Hence we limited the term size to fewer than 1000 genes and
only examined those that passed p-value threshold for FDR correction.


An ORA was the performed on the up-regulated significant differentially expressed genes:
```{r gost_up, message=FALSE, warning=FALSE, echo=FALSE}
gostres_up <- gostQuery(genelist = upregulated_genes)
knitr::kable(summariseGost(gostres = gostres_up))
```


An interactive Manhattan plot that summarises the query result of the up-regulated top-hits:
```{r, warning=FALSE, message=FALSE, echo=FALSE}
gprofiler2::gostplot(gostres = gostres_up)
```


Summaries of the ORA result for the down-regulated genes:
```{r gost_down, message=FALSE, warning=FALSE, echo=FALSE}
gostres_down <- gostQuery(genelist = downregulated_genes)
knitr::kable(summariseGost(gostres = gostres_down))
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
gprofiler2::gostplot(gostres = gostres_down)
```

The result of the ORA using all the significant differentially expressed genes was compared against the other two using the up-regulated set and the down-regulated set in terms of the pathways returned from each source of annotation data.
The number of gene sets returned appears to be proportional to the length of the query gene list. All these three results using three different gene lists reveal different information.
In terms of GO biological processes, the result for all differentially expressed genes involve regulation of *neutrophil*, *leukocyte*, and *myeloid cells*. The result for up-regulated genes does not seem to have many informative GO terms except for viral gene expression and viral transcription. The result for the down-regulated genes has terms involve *type I interferon* and defense response to virus.
In terms of REACTOME, pathways involving *neutrophil* regulation and *myeloid leukocyte* also appear in the result for all differentially expressed genes. The result for the up-regulated genes does not seem to have many informative terms either. The result for the down-regulated genes also shows signaling pathways for *interferon*.
In terms of wikipathways, the results for all differentially expressed genes shows *T-Cell antigen Receptor (TCR) Signaling Pathway* and *Type II interferon signaling (IFN)*; pathways involving type I interferon appear too. The result for the up-regulated genes also have pathways involving *TCR*. The result for the down-regulated genes is similar to the previous two sources, showing pathways involving *Type I interferon* and *Type II interferon*.
In terms of KEGG, the result for all differentially expressed genes has "Coronavirus disease - COVID-19" has the top hit pathway, expected but not informative; many cancer-relative pathways also appear. The result for the up-regulated genes shows a pathway involving TCR. The KEGG pathways for the down-regulated genes are mostly cancer-related.

The ORA results highly support the conclusions drawn in the original paper.
The authors also found pathways such as neutrophil activation, innate immune response,
immune response to viral infection, type-I interferon signaling,
and cytokine production for 771 upregulated genes they discovered in COVID-19 patients;
pathways they found for 1231 down-regulated genes include
lymphocyte differentiation and T-cell activation and regulation,
which also appear in our ORA result. The authors concluded that
T-cell are suppressed while neutrophils are activated as
an indicator of host response to COVID-19 represented in the transcriptomic changes.[@GSE152641]

There have been prior researches on host immune responses to COVID-19 showing findings that support these ORA results.
An immune analysis on COVID-19 patients by Hadjadj et al. has shown findings that coincide with our ORA result.
Their multiplex gene expression analysis showed
an up-regulation of genes involved in type I interferon signaling in COVID-19 patients[@Hadjadj718],
while pathways involving type I interferon are present in our ORA results for down-regulated genes in healthy control.
Anurag et al. observed increased total leukocyte count and differential neutrophil count in patients with severe COVID-19 infection[@anurag2020],
and GO terms *leukocyte activation involved in immune response* and *neutrophil activation involved in immune response* are present in the ORA result for all differentially expressed genes.
Kalfaoglu B. et al. discovered a unique dynamics of T-cells in severe COVID-19 patients: T-cells become hyperactivated, proliferate and die rapidly before differentiating into Treg, while they have shown that the majority of Treg-type genes are regulated by TCR signaling[@hyper_Tcell]; this supports the involvement of T-cell signaling pathways in our ORA result.
